---
title: "Bachelor_thesis"
author: "Luis"
date: "2026-06-14"
output: html_document
---

This R markdown document is part of the bachelor thesis "On Why Scoring Functions Cannot Assess Tail Properties: A More Detailed Analysis". All derivations and explanations can be found there.

## Storing the distributions

A distribution $G \in \mathcal{F}_\text{exp}$ is stored as a data frame with entries lambda and theta. lambda is a vector of the $\lambda_i$ which make up the convex combination. theta is a vector of the $\theta_i$ which are the rates of the "pure" exponential distributions of $G$.

We take G to be the "true" distribution, i.e. the distribution our realisations come from. F is a distribution with heavier tail as requested in Theorem 4.4.

```{r distributions}
G <- data.frame(
  lambda <- c(0.2,0.3,0.5),
  theta<- c(1,3,10)
)

F <- data.frame(
  lambda <- c(1),
  theta <- c(0.5)
)
```

A function that gives us a convex combination of the distributions

```{r convex combination of two distributions}
convex_combine <- function(F,G, lambda_conv_comb){
  # define new distribution by the convex combination of F and G
  F_lambda <- data.frame(
    lambda <- c(lambda_conv_comb*F$lambda, (1-lambda_conv_comb)*G$lambda),
    theta <- c(F$theta, G$theta)
  )
  colnames(F_lambda) <- c("lambda", "theta")
  
  return(F_lambda)
}
```

## Sampling

Get a sample of size $n$ from the distribution $G \in \mathcal{F}_\text{exp}$

```{r sampling from G}
get_sample <- function(G, n){
  
  l <- length(G$lambda)
  
  # initialise the result vector
  result <- c()
  
  # Similar to Remark 3.1, sampling from a convex combination of 
  # exponential distributions is like choosing which exponential distribution 
  # to choose from via a multinomial distribution
  pick <- rmultinom(1, size = n, prob= G$lambda)
  
  for (i in 1:l){
    # sample as many realisations from the corresponding distribution
    # and append to the result vector
    result <- c(result, rexp(pick[i], rate = G$theta[i]))
  }
  return (result)
}
```

## CRPS and the CRPS_bar approximation

The formula to calculate the CRPS for $G \in \mathcal{F}_\text{exp}$ and $y \in [0,\infty)$ $$\text{CRPS}(G,y) = y + \sum_{i=1}^n \frac{2\lambda_i}{\theta_i}(e^{-\theta_i y} - 1) +  \sum_{i,j=1}^n \frac{\lambda_i\lambda_j}{\theta_i+\theta_j} = y+ \sum_{i=1}^n \left( \frac{2\lambda_i}{\theta_i}(e^{-\theta_i y} - 1) +  \sum_{j=1}^n \frac{\lambda_i\lambda_j}{\theta_i+\theta_j}\right) $$

```{r CRPS}
CRPS <- function(G,y){
  result <- y
  for (i in 1:length(G$lambda)){
    result <- result + 2*G$lambda[i]/G$theta[i]*(exp(-G$theta[i]*y)-1)
    
    # the double sum term
    for (j in 1:length(G$lambda)){
      result <- result + G$lambda[i]*G$lambda[j]/(G$theta[i]+G$theta[j])
    }
  }
  return(result)
}
```

Estimate $$\overline{\text{CRPS}}(F,G) \approx \frac{1}{n}\sum_{i=1}^n \text{CRPS}(F,y_i) \text{ where the } y_i \text{ are i.i.d. realisations of } Y \sim G$$

```{r CRPS approximation}
# Estimate CRPS_bar(F,G) using n i.i.d realisations
CRPS_bar_hat <- function(F,G, n=1000000){
  sample <- get_sample(G, n)
  score <- mean(CRPS(F, sample))
  return(score)
}
```

And a version to apply to a list of distributions

```{r vectorised CRPS_bar_hat}
# From a list of distributions, get a vector of their estimated CRPS_bar
CRPS_bar_hat_vec <- function(list_distr, G_true, n=1000000){
  #initialise the result vector
  result <- c()
  
  for (i in list_distr){
    result <- c(result, CRPS_bar_hat(i, G_true, n))
  }
  
  return(result)
}
```

## Probability density function

Get the probability density function of a distribution

```{r pdf}
# Higher order function taking a distribution as input and
# returning the pdf of this distribution
pdf <- function(G){
  G_pdf <- function(x){
    result <- 0
    for (i in 1:length(G$lambda)){
      result <- result + G$lambda[i]*G$theta[i]*exp(-G$theta[i]*x)
    }
    return(result)
  }
  return(G_pdf)
}
```

Get the logarithm of the probability density function of a distribution

```{r log(pdf)}
# Higher order function taking a distribution as input and
# returning the logarithm of the pdf of this distribution
log_pdf <- function(G){
  G_log_pdf <- function(x){
    result <- 0
    for (i in 1:length(G$lambda)){
      result <- result + G$lambda[i]*G$theta[i]*exp(-G$theta[i]*x)
    }
    return(log(result))
  }
  return(G_log_pdf)
}
```

## Plotting

To visually compare the distributions, we plot their pdfs

```{r plot}
# Plots the pdfs of a list of distributions in one plot
plot_distr <-  function(list_of_distributions){
  
  # get first distribution back as data frame
  distr_1 <- as.data.frame(list_of_distributions[[1]])
  colnames(distr_1) <- c("lambda", "theta")
  # take its pdf
  pdf_distr_1 <- pdf(distr_1)
  # plot the first distribution
  curve(pdf_distr_1, xlim=c(0,3), ylim = c(0,1.5), ylab = "pdf", col=1)
  
  # plot the remaining pdfs in case there are any
  if (length(list_of_distributions) > 1){
    for (i in 2:length(list_of_distributions)){
      # get i-th distribution back as data frame
      distr_i <- as.data.frame(list_of_distributions[i])
      colnames(distr_i) <- c("lambda", "theta")
      # take its pdf
      pdf_distr_i <- pdf(distr_i)
      # plot the i-th distribution
      curve(pdf_distr_i, add = TRUE, col = i)
    }
  }
  
  # add title if needed
  # title("Probability density function of the considered distributions")
  
  # prepare legend
  lege <- names(list_of_distributions)
  if (length(lege) == 0){
    lege <- 1:length(list_of_distributions)
  }
  # add legend
  legend("topright", legend=lege, col=1:length(list_of_distributions), lty=1, lwd =2,
         ncol = ceiling(length(list_of_distributions)/5) , cex = 0.75)
}
```

To visually compare the tails of the distributions, we plot the log of their pdfs

```{r log plot}
# Plots the logarithm of the pdfs of a list of distributions in one plot
plot_log_distr <-  function(list_of_distributions){
  
  # get first distribution back as data frame
  distr_1 <- as.data.frame(list_of_distributions[[1]])
  colnames(distr_1) <- c("lambda", "theta")
  # take its log(pdf)
  log_pdf_distr_1 <- log_pdf(distr_1)
  # plot the first distribution
  curve(log_pdf_distr_1, xlim=c(0,3), ylim = c(-3,0), ylab = "log(pdf)", col=1)
  
  # plot the remaining log(pdf)s in case there are any
  if (length(list_of_distributions) > 1){
    for (i in 2:length(list_of_distributions)){
      # get i-th distribution back as data frame
      distr_i <- as.data.frame(list_of_distributions[i])
      colnames(distr_i) <- c("lambda", "theta")
      # take its log(pdf)
      log_pdf_distr_i <- log_pdf(distr_i)
      # plot the i-th distribution
      curve(log_pdf_distr_i, add = TRUE, col = i)
    }
  }
  
  # add title if needed
  # title("Logarithm of the pdf of the considered distributions")
  
  # prepare legend
  lege <- names(list_of_distributions)
  if (length(lege) == 0){
    lege <- 1:length(list_of_distributions)
  }
  # add legend
    legend("topright", legend=lege, col=1:length(list_of_distributions), lty=1, lwd =2, 
            ncol = ceiling(length(list_of_distributions)/5), cex = 0.75)
}
```

## Illustration of Theorem 4.4

First, define the distributions we want to work with

```{r define distributions}
F_0.9   <- convex_combine(F, G, 0.9)
F_0.8   <- convex_combine(F, G, 0.8)
F_0.7   <- convex_combine(F, G, 0.7)
F_0.6   <- convex_combine(F, G, 0.6)
F_0.5   <- convex_combine(F, G, 0.5)
F_0.4   <- convex_combine(F, G, 0.4)
F_0.3   <- convex_combine(F, G, 0.3)
F_0.2   <- convex_combine(F, G, 0.2)
F_0.15  <- convex_combine(F, G, 0.15)
F_0.1   <- convex_combine(F, G, 0.1)
F_0.05  <- convex_combine(F, G, 0.05)
F_0.04  <- convex_combine(F, G, 0.04)
F_0.03  <- convex_combine(F, G, 0.03)
F_0.02  <- convex_combine(F, G, 0.02)
F_0.01  <- convex_combine(F, G, 0.01)
F_0.005 <- convex_combine(F, G, 0.005)
```

Make a list of the distributions

```{r make list}
distributions <- list(
  "l = 1"     = F,
  "l = 0.9"   = F_0.9,
  "l = 0.8"   = F_0.8,
  "l = 0.7"   = F_0.7,
  "l = 0.6"   = F_0.6,
  "l = 0.5"   = F_0.5,
  "l = 0.4"   = F_0.4,
  "l = 0.3"   = F_0.3,
  "l = 0.2"   = F_0.2,
  "l = 0.15"  = F_0.15,
  "l = 0.1"   = F_0.1,
  "l = 0.05"  = F_0.05,
  "l = 0.04"  = F_0.04,
  "l = 0.03"  = F_0.03,
  "l = 0.02"  = F_0.02,
  "l = 0.01"  = F_0.01,
  "l = 0.005" = F_0.005,
  "l = 0"     = G
)

# Define a vector of the lambdas used for plotting
lambda_vector <- c(1,0.9,0.8,0.7,0.6,0.5,0.4,0.3,0.2,0.15,
                   0.1,0.05,0.04,0.03,0.02,0.01,0.005,0)
```

Plot the pdfs and the log(pdf)s of the distributions

```{r}
plot_distr(distributions)
title("Probability density function of F_l = l*F + (1-l)*G")
plot_log_distr(distributions)
title("Logarithm of the probability density function of F_l = l*F + (1-l)*G")
```

We calculate the estimation

```{r calculate estimation}
# number of observations we want to compare the distributions to
n_obs <- 100000000

CRPS_bar_hat_vector <- CRPS_bar_hat_vec(distributions, G, n_obs)
```

Plot the estimation and close ups of the interesting regions

```{r plot the estimated CRPS}
plot(lambda_vector, CRPS_bar_hat_vector, 
     ylab = "Estimated CRPS_bar(F_l, G)", xlab=("l"), pch = 4)
title("Estimated CRPS_bar of l*F + (1-l)*G")

plot(lambda_vector, CRPS_bar_hat_vector, 
     ylab = "Estimated CRPS_bar(F_l, G)", xlab=("l"), 
     xlim=c(0,0.2), ylim = c(0.22,0.26),pch = 4)
# add a line to better differentiate
abline(h = CRPS_bar_hat_vector[length(CRPS_bar_hat_vector)])
title("Estimated CRPS_bar of l*F + (1-l)*G")

plot(lambda_vector, CRPS_bar_hat_vector, 
     ylab = "Estimated CRPS_bar(F_l, G)", xlab=("l"), 
     xlim=c(0,0.05), ylim = c(0.23,0.235),pch = 4)
# add a line to better differentiate
abline(h = CRPS_bar_hat_vector[length(CRPS_bar_hat_vector)])
title("Estimated CRPS_bar of l*F + (1-l)*G")
```
